---
title: "Open Canada Vital Statistics Tracker"
github-repo: open-canada/vitals
url: https\://open-canada.github.io/Apps/vitals
license: MIT 
favicon: "rCanada.ico"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    social: menu
    source_code: https://github.com/open-canada/vitals
    vertical_layout: fill # scroll #fill # 
runtime: shiny
---

```{r 00_vital_meta.Rmd, child = '00_vital_meta.Rmd'}
```

```{r -globals, include=F}
knitr::opts_chunk$set(echo = F)
source ("01_vitals_ca_LTE.R")

if (F) {
  dtCached 
  dtGeo
  dtVac
  dtVacAllAgesAllSexes
  choicesGEO
  choicesCauses
  dateToday
  dateMax
  dateCached
}

if (T) {
  # For faster/memory efficient processing declare these as global 
  # in OOP, these (below and the ones above and default input values) would be part of the class CVitals, 
  dt <- data.table()
  dtAll <- data.table()
  dt0 <- data.table()
  region1 <- ""
  cause1 <- ""
}

# * 4.a Set parameters: static ----

in0 <- list(
  state = choicesGEO %wo% c("Yukon", "Northwest Territories", "Nunavut", "Newfoundland and Labrador", "Prince Edward Island" ),
  # state =  c("Canada", "Quebec" , "Ontario", "Alberta", "British Columbia" ), 
  cause = choicesCauses[c(1,2,5, 12, 15:16)], # 5:9, 7, 
  # clump_nonselected=T,
  
  read_from="Cached Data",
  average = F,
  alternative_view=F,
  per_million = T,
  keep_scale = F,
  lm=F,
  se=F,
  
  vaccination=F,
  vax="Total",
  # dose=2,
  
  date = c("2019-12-01", as.character(dateToday))
)

if ( !shiny::isRunning() ) {  
  cat("Relax - Shiny is NOT Running :)")
  input <- in0
}

# Does not work yet :()
# This would have been neat idea - so that the same code can be used with reactive (shiny) AND static (non-shiny) execution
# Should ask stackoverfloor
.reactive <- function(A) {
  if (!shiny::isRunning()) {
    A
  }  else reactive(A)
}
```

# Search criteria: {.sidebar data-width=210}

<font size=-1>

Data Source: [Statistics Canada Table 13-10-0810-01](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310081001)

```{r shiny_input}

dateCached = ymd("2022-01-10")
# dateReleased = ymd("2021-12-06"); 
dateReleased = ymd("2022-01-10"); 
# dateCached = dateReleased

renderText(paste0("Data loaded on: ", ifelse(input$read_from=="Cached Data" , dateCached %>% as.character(), dateToday%>% as.character() )))
renderText(paste0("Last record in data: ", 
                  # dateMax
                  r.dtAll()$Date %>% max(na.rm = T)
                  )
           )

# 4.b Set parameters: interactive ----

radioButtons("read_from", NULL, # "Read data from ",
             choices = c("Use Live", "Cached Data" ),
             selected = "Cached Data" , inline=T)

checkboxInput("alternative_view", "Reverse axes",  in0$alternative_view) 

checkboxInput("per_million", "Show per million",  in0$per_million)
checkboxInput("average", "Average over three weeks",  in0$average)
checkboxInput("keep_scale", "Keep the same scale",  F)
checkboxInput("vline", "Overlay vertical markers",  F)

checkboxInput("lm", "Overlay trendline",  F)
renderUI({
  if (input$lm){
    checkboxInput("se", "Show confidence interval",  F)
  }
})

checkboxInput("vaccination", "Show vaccination rates ...",  in0$vaccination)
renderUI({
  if (input$vaccination){
    radioButtons("vax", NULL, c("Total", "Weekly"), inline=T)
  }
})
# renderUI({
#   if (input$vaccination){
#         # radioButtons("doze", "... for", c("1+", "2+), inline=T)
#     sliderInput("doze", "Dose number to show:", min=1, max=4)
#   }
# })

# renderUI({
#   if (input$vaccination){
#     sliderInput("vax_rate_line", "Vaccination rate to mark:", min=0, max=100, value=3, step=1, ticks=F, width="100%")
#   }
# })


selectInput('state', h5("Select region(s):"), multiple=T, choices = choicesGEO, selected = in0$state, selectize=F, size=5 )
selectInput('cause', h5("Select death cause(s):"), multiple=T, choices = choicesCauses, selected = in0$cause, selectize=F, size=5    )


# Predefine some periods of interest:

period = c(
  ymd("2010-01-05"), # early prepandemic, early flu waves 
  # ymd("2016-01-01"), # late pre-pandemic, later flu  waves
  # ymd("2020-01-01"), # four new cause categories added: 3 for covid + NO_INFO
  ymd("2020-03-24"), # early pre-vax pandemic, first COVID death
  ymd("2020-08-01"), # late pre-vax pandemic, end of first wave
  ymd("2020-12-14"), # early vax, first vax shot 
  ymd("2021-05-01"), # late vax, vax rate >= 3%
  # ymd("2021-12-02"), # 3rd doses (first booster)
  # ymd("2021-10-04")     # dateToday dateMax   
  dateToday
)


# Somehow does not work...
# selectInput(
#   "period",  h5("Select date range"),  period,
#   selected = period[2:length(period)] %>% as.character,
#   selectize=F, size=5,
#   multiple = T
# )
# renderUI({
  # dateRangeInput("date", h5("... adjust range"), weekstart=1, 
  #                start = ymd(input$period) %>% min(na.rm=T),
  #                end = ymd(input$period) %>% max(na.rm=T)
  #                )
  # in0$date[1] = input$period %>% min
  # in0$date[2] = input$period %>% max
  # dateRangeInput("date", h5("Select date range"), weekstart=1, start = in0$date[1], end = in0$date[2])
# })


dateRangeInput("date", h5("Select date range"), weekstart=1, start = in0$date[1], end = in0$date[2])

```


</font>



```{r *r.dtAll r.dtVAc}

r.dtVac0  <- reactive({
  if (input$vax=="Weekly"){ 
    dtVacAllAgesAllSexes[, dose1_rate := dose1_rate - shift(dose1_rate, 1), by = .(GEO)]
    dtVacAllAgesAllSexes[, dose2_rate := dose2_rate - shift(dose2_rate, 1), by = .(GEO)]
  } else{
    dtVacAllAgesAllSexes
  }
})

r.dtAll  <- reactive({
  
  if (input$read_from=="Cached Data") {
    dt = copy(dtCached)
  } else {
    # req(r.dtFile1())
    # req(input$read_from)
    # r.dtFile1()  %>% select(input$columns1)   %>%
    #          mutate_if(is.factor, as.character) 
    
    dt <- cansim::get_cansim("13-10-0810-01") %>%  setDT(dt)
    # Quick view of the data
    dt
    dt %>% names 
    # Remove unneeded columns and simplify values
    # dt[, (names(dt)[c(1,3:20,24)]):=NULL]
    dt <- dt[, c("Date", "GEO", "val_norm", "Cause of death (ICD-10)")]
    dt[, Date := ymd(Date)]; 
    dateMax <- dt$Date %>% max; dateMax # "2021-10-02"
    dateLive <- dateToday
    dt$GEO %>% unique() %>% sort
    dt[, GEO := gsub(", place of occurrence", "", GEO)]
    
    # dt <- dtGeo[dt, on="GEO"]
    dt[, GEO:=fct_relevel(GEO, choicesGEO)]
    setnames(dt, "val_norm", "value")
    setcolorder(dt, c("Date",  "GEO", "Cause of death (ICD-10)", "value"))
  }
  
  # NB: anomalies were population is less than million.
  
  if (input$per_million) {
    dt <- dtGeo[dtCached, on="GEO"]
    dt [, GEO:=fct_relevel(GEO, choicesGEO)]
    dt [, value:=round(1000000*value/population)]
    setcolorder(dt, c("Date",  "GEO", "population", "Cause of death (ICD-10)", "value"))
  } 
  
  if (input$average) {
    convolution_window <- 3
    dt[, value := frollmean(value, convolution_window, align = "right", fill = 0), by = .(GEO, `Cause of death (ICD-10)`)]
  }
  
  if (input$vaccination) {
    # dt <- dtVacAllAgesAllSexes[dt, on=.(prename=GEO, week_end=Date)]
    dt <- r.dtVac0()[dt, on=c("GEO", "Date")]
  }
  
  dt[, GEO:=fct_relevel(GEO, choicesGEO)]
  # setcolorder(dt, c("Date", "GEO"))
  dtAll = dt 
})
```

```{r *r.dt0 r.dtSummary}

r.dt0  <- reactive({
  
  q <- quote(Date >= input$date[1] &  Date <= input$date[2] & GEO %in% input$state & as.character(`Cause of death (ICD-10)`) %in% input$cause )
  
  # dt0 <<- dtAll[eval(q)]
  dt0 <<- r.dtAll()[eval(q)]
})

r.dtSummary  <- reactive({
  dt0 <<- r.dt0()
  
  dtS <- dt0 [ , .(
    `Weekly lowest`=min(value, na.rm = T) %>% round(0),    
    `Mean`=mean(value, na.rm = T) %>% round(1),   
    `SD`=sd(value, na.rm = T) %>% round(1), 
    `Weekly highest`=max(value, na.rm = T) %>% round(0),
    
    # LM = list(list(lm(value ~ as.integer( Date - min(Date))),  na.action=na.omit)),
    `Weekly dynamics`=( lm(value ~ as.integer( Date - min(Date)),  na.action=na.omit)$coefficients[2] * 7 ) %>% round(1)
  ),    by=.(`Cause of death (ICD-10)`,GEO)]
  
  dtS
})

# Smoothing method (function) to use, accepts either NULL or a character vector, e.g. "lm", "glm", "gam", "loess" or a function, e.g. MASS::rlm or mgcv::gam, stats::lm, or stats::loess. "auto" is also accepted for backwards compatibility. It is equivalent to NULL.
# 
# For method = NULL the smoothing method is chosen based on the size of the largest group (across all panels). stats::loess() is used for less than 1,000 observations; otherwise mgcv::gam() is used with formula = y ~ s(x, bs = "cs") with method = "REML". Somewhat anecdotally, loess gives a better appearance, but is O(N^2) in memory, so does not work for larger datasets.
# 
# If you have fewer than 1,000 observations but want to use the same gam() model that method = NULL would use, then set method = "gam", formula = y ~ s(x, bs = "cs").
# 
# formula	
# Formula to use in smoothing function, eg. y ~ x, y ~ poly(x, 2), y ~ log(x). NULL by default, in which case method = NULL implies formula = y ~ x when there are fewer than 1,000 observations and formula = y ~ s(x, bs = "cs") otherwise.

```

# Deaths 

<!-- # Deaths by Cause -->

## ........ {.tabset .tabset-fade }




### Overview
<!-- ### Multi-dimensional plot -->

```{r r.g, out.height="100%" }

# ```{r plotOverview}
r.g <- reactive({

  # https://stackoverflow.com/questions/70656074/showing-something-my-text-instead-of-error-message-text-while-data-is-being-load
    # req(input$)
  
  dt0 <<- r.dt0()
  # if (is.null(dt0)) {
  #   return ggplot()
  # }
  
  dt0$vaccination <- factor("Vaccination", levels=c("Vaccination", "Dummy") )
  
  g2 <-  ggplot(dt0) +   guides(col="none") +
    # scale_x_date(guide = guide_axis(n.dodge = 2)) +
    geom_line(aes(Date, dose2_rate)) + 
    geom_line(aes(Date, dose1_rate), linetype=3) +
    labs(title = NULL,  x = NULL,   y = "(%)",
         caption = "Source: https://health-infobase.canada.ca/covid-19/vaccination-coverage")

  
  g <-  ggplot(dt0) +  
    theme(legend.position = "bottom") +
    geom_line(aes(Date, value, col = `Cause of death (ICD-10)`)) +
    # guides(x =  guide_axis(angle = 90)) +
    # scale_x_date(guide = guide_axis(n.dodge = 2)) +
    # scale_x_discrete(guide = guide_axis(angle = 90)) +
    # theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs( title = NULL, x = NULL,  
          y = ifelse(input$per_million,  "Deaths per million per week", "Deaths per week"), 
          caption = "Source: Statistics Canada - Table 13-10-0810-01"     )
  
  
  if (input$lm==T) {
    
    if (input$vline) {  
      for (i in 1:(length(period)-1)) {
        g <- g + geom_smooth(aes(Date, value), method="lm", se = input$se, na.rm=T, alpha=0.5, linetype=3, size=0.5, col="black", data=dt0[Date>=period[i] & Date<period[i+1]])
      }
    } else {
      g <- g + 
        geom_smooth(aes(Date, value), method="lm", se = input$se, na.rm=T, alpha=0.5, linetype=3, size=0.5, col="black") 
    }
  }
  
  
  if (input$vline) {  
    alp = 0.2  
    for (i in 1:length(period)) {
      g <- g + geom_vline(xintercept=period[i], linetype=5, aplha=alp,  col="grey")
      g2 <- g2 + geom_vline(xintercept=period[i], linetype=5, aplha=alp,  col="grey")
    }
    
    g <- g + xlim(input$date[1], input$date[2])
    g2 <- g2 + xlim(input$date[1], input$date[2])
  }
  
  # if (F) { # should be by province!
  #   g <- g+ geom_vline(xintersept=dt0[as.integer(dose2_rate) == input$vline_doze]$Date %>% median (na.rm = T))
  # }
  

  if (input$alternative_view == F) {

    # VERTICAL LAYOUT ############################################### #
    
    g <- g +  
      facet_grid(`Cause of death (ICD-10)` ~ GEO ,  scales = ifelse (input$keep_scale, "fixed", "free")  ) 

    if (input$vaccination==T) {
      g2 <- g2 + facet_grid( vaccination ~ GEO ) 
      # a <- max(1, input$cause %>% length - 4)
      a=5
      g <- ggpubr::ggarrange(g2, g, nrow = 2, heights = c(1,a))
    } 
    # else {
    #   g <- g1
    # }
    
  } else {
    
  # HORIZONATAL LAYOUT ############################################# #
    
    g <- g + 
      guides(col="none")  + 
      scale_x_date(guide = guide_axis(n.dodge = 2)) +
      facet_grid(GEO ~ `Cause of death (ICD-10)`, 
                 scales = ifelse (input$keep_scale, "fixed", "free") ) 
    if (input$lm==T) {
      g <- g + geom_smooth(aes(Date, value), method="lm", se = F, na.rm=T, alpha=0.5, linetype=4, size=0.3, col="black") 
    }

    if (input$vaccination==T) {
  
      g2 <-  g2 +   
        guides(col="none") +
        scale_x_date(guide = guide_axis(n.dodge = 2)) +
        facet_grid(GEO ~ vaccination , scales = "free") +
        labs(      title = NULL,         x = NULL,        y = "Vaccination rate (%)", 
                   caption = "https://health-infobase.canada.ca"  )
      
      g <- ggpubr::ggarrange(g, g2, ncol = 2, widths=c(5,1))
      
    } 
    
  }
  g
})


renderPlot( r.g() )


```



<!-- ###  Interactive time-series  -->

<!-- Plot shows time-series for all causes in  one region  (the first in the list of the selected regions).    -->

### By cause

```{r dygraphs}

# See: https://rstudio.github.io/dygraphs/

dygraphs::renderDygraph ({


  dt0 <- r.dt0()
  region1 <<- dt0$GEO %>% unique %>% .[1]
  
  dt0_wideGeo1 <- dt0[GEO == region1, !c("GEO"), with = T] %>% dcast( Date ~ `Cause of death (ICD-10)`, value.var = "value")
  
  
  # setcolorder(dt0_wideGeo1, "Date"); 
  dts  <- as.xts.data.table(dt0_wideGeo1)
  
  # dy <- dygraph.title( dts, region1)
  
  dy <- dygraph(dts, main = region1) %>%
    dyOptions(fillGraph = F, stepPlot = F, drawGrid = T, drawPoints = TRUE, pointSize = 2) %>%
    dyHighlight(highlightCircleSize = 5,highlightSeriesBackgroundAlpha = 0.2, hideOnMouseOut = F) %>%
    dyRangeSelector()  %>%
    dyAxis("y", label=ifelse(input$per_million,  "Deaths in million / week", "Deaths / week")) 
  
  # Does not work yet ...
  # We can also add second axis 
  # if (input$vaccination){
  #   dts1 <- as.xts.data.table(r.dtVac0()[GEO==region1, !c("GEO"), with = T])
  #   dy <- dy %>% dySeries(dts1, axis = "y2" )
  # }

  dy
})
```

### By province

```{r}

dygraphs::renderDygraph ({
  
  dt0 <- r.dt0()
  cause1 <<- dt0$`Cause of death (ICD-10)`%>% unique %>% .[1]
  
  
  dt0_wideCause1 <- dt0[`Cause of death (ICD-10)`==cause1, !c("Cause of death (ICD-10)"), with=T] %>% 
    dcast(Date ~ GEO, value.var="value")
  
  dts  <- as.xts.data.table(dt0_wideCause1)
  
  # dy <- dygraph.title( dts, cause1)
  dy <- dygraph(dts, main = cause1) %>%
    # dySeries(input$var1, color = input$color1, strokePattern = input$stroke1,  axis = input$axis1 )  %>%
    dyOptions(fillGraph = F, stepPlot = F, drawGrid = T, drawPoints = TRUE, pointSize = 2) %>%
    dyHighlight(highlightCircleSize = 5,highlightSeriesBackgroundAlpha = 0.2, hideOnMouseOut = FALSE) %>%
    dyAxis("y", label=ifelse(input$per_million,  "Deaths in million / week", "Deaths / week")) %>%
    # dyAnnotation("2021-6-10", text = "10%", tooltip = "Fully Vaccinated Rate 10%") %>%
    # dyAnnotation("2021-5-1", text = "3%", tooltip = "Fully Vaccinated Rate 3%") %>%
    # dyAnnotation("2021-4-6", text = "2%", tooltip = "Fully Vaccinated Rate 2%") %>%
    # dyAnnotation("2021-2-18", text = "1%", tooltip = "Fully Vaccinated Rate 1%") %>%
    dyRangeSelector()
  
  
  # dy <- dy %>% 
  #   dyAxis("y", label=ifelse(input$per_million,  "Deaths in million / week", "Deaths / week")) 
  
# 
#   if (input$vaccination){
#     dy <- dy %>% 
#       dySeries(dcast(dtVacAllAgesAllSexes, Date ~ GEO, value.var="dose2_rate"), color = "red", axis = "y2", strokePattern=2 )  %>% 
#       dySeries(dcast(dtVacAllAgesAllSexes, Date ~ GEO, value.var="dose1_rate"), color = "red", axis = "y2", strokePattern=2 ) 
#   }
  
  dy
})

```


### By cause (alt.) 


```{r ggplotly}

# See: 
# https://plotly.com/r/, https://plotly.com/ggplot2/getting-started/
# https://plotly-r.com/

plotly::renderPlotly( {
  
  dt0 <<- r.dt0()
  # Interactive plot can visualize only data for one region
  region1 <- dt0$GEO %>% unique %>% .[1]
  g <-ggplot(dt0[GEO==region1]) +
    geom_line(aes(Date, value,   col = `Cause of death (ICD-10)`)) +
    labs(        
      title = paste0(region1)  ,         
      x = NULL,
      y = ifelse(input$per_million,  "Deaths in million / week", "Deaths / week")  
    )
  
  if (input$vaccination==T) {
    g <-  g +
      geom_line(aes(Date, dose2_rate)) +
      geom_line(aes(Date, dose1_rate), linetype=3) +
      labs(        
        title = region1  ,         
        x = NULL,        
        y = ifelse(input$per_million,  "Vaccinatrion rate (%) | Deaths in million per week", "Vaccinatrion rate (%) | Deaths per week")  
      )
  }
  
  plotly::ggplotly( g ) 
})

```


### By province (alt.)



```{r }

plotly::renderPlotly( {
  
  dt0 <- r.dt0()
  cause1 <- dt0$`Cause of death (ICD-10)`%>% unique %>% .[1] 
  # Interactive plot can visualize only data for one cause
  g <-
    dt0[`Cause of death (ICD-10)` == cause1] %>% 
    ggplot() +
    geom_line(aes(Date, value,  col = GEO)) +
    labs(        
      title = paste0(cause1)  ,         
      x = NULL,        
      y = ifelse(input$per_million,  "Deaths in million / week", "Deaths / week") 
    )
  
  if (input$vaccination==T) {
    g <-  g +
      geom_line(aes(Date, dose2_rate, col = GEO)) +
      geom_line(aes(Date, dose1_rate, col = GEO), linetype=3) +
      labs(        
        title = cause1,         
        x = NULL,        
        y = ifelse(input$per_million,  "Vaccinatrion rate (%) | Deaths in million per week", "Vaccinatrion rate (%) | Deaths per week")   
      )
  }
  
  plotly::ggplotly( g ) 
})



```



###  Summary 




```{r datatable}
DT::renderDataTable(
  r.dtSummary() %>% datatable.title()
)

```



###  Full table


```{r}
# See: https://rstudio.github.io/DT/

DT::renderDataTable({
  
  dt0 <- r.dt0() 
  
  if (input$vaccination) {
    dt0.wide <-  dt0 %>% dcast(GEO    +   Date + dose1_rate + dose2_rate ~ `Cause of death (ICD-10)`, value.var = "value")
  } else {
    dt0.wide <- dt0 %>% dcast(GEO    +   Date ~ `Cause of death (ICD-10)`, value.var = "value")
  }
  
  if (input$per_million)
    dt0.wide <- dtGeo[dt0.wide, on="GEO"][order(GEO, -Date)]
  
  dt0.wide %>% datatable.title()
})


```


# Births

## .....

### UNDER CONSTRUCTION

A place-holder for visualizing  [Other related Canadian Vital Statistics data]

Note that most  databases listed in [Other related Canadian Vital Statistics data] are updated annually and hence cannot be used for weekly time-series analysis enabled by this App.

<!-- and/or provide less granular data than the ones that are currently visualized by this App, and hence  -->




<!-- # Deaths by Age -->

<!-- ## ........ {.tabset .tabset-fade } -->

<!-- ### Plot -->

<!-- UNDER DEVELOPMENT -->


```{r eval=FALSE, include=FALSE}
# 5. Open Deaths by Age ----


dtAge <- readRDS(paste0(id, ".Rds")) %>% setDT
if ( is.na(dtAge[.N]$GEO) ) {
  dateCached <- dtAge[.N]$Date %>% ymd
  dtAge <- dtAge[1:(.N-1)]
}

setnames(dtAge, old=c("Age at time of death", "Sex"), new=c("age", "sex"))

```






```{r 99_vital_info.Rmd, child = '99_vital_info.Rmd'}
```



```{r STOP_HERE, include=FALSE}
knitr::knit_exit()
```
Any other codes/text, which you dont want to be shown (yet) in the complied App, can be placed here. For example debugging pane or extra statistical analysis tab




# Debug 

## ...

### Debug window



#### Input
```{r print input1}
# renderPrint( paste0("Shiny is Running: ", shiny::isRunning() ) )
renderPrint( input )
renderPrint( input$date )
renderPrint( input$period )
renderPrint( ymd(input$period) )
renderPrint( ymd(input$period) %>% min)
renderPrint( ymd(input$period) %>% max)
renderPrint( paste0("region1 ", region1))
renderPrint( paste0("cause1 ", cause1))

# cat("r.dtAll()")
# renderPrint( r.dtAll() %>% print )
cat("r.dt0()")
renderPrint( r.dt0() %>% print )
cat("r.dt0_wideGeo1()")
renderPrint( r.dt0_wideGeo1() %>% print )
```




