---
title: "Open Canada Data"
github-repo: open-canada/vitals
url: https\://o-canada.shinyapps.io/vitals/data.Rmd
license: MIT 
# favicon: "R4GC.ico"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    social: menu
    source_code: https://github.com/open-canada/vitals
    vertical_layout: fill # scroll 
runtime: shiny
---


```{r -globals, include=F}
knitr::opts_chunk$set(echo = F)
source ("00_common.R")
library(dplyr)
library(ckanr)
library(cansim)

```


# Search criteria: {.sidebar data-width=210}

```{r}

# selectInput
radioButtons("data.source", "Choose data origin:",
             choices = c("Upload file", "URL",
                         "StatCan Table", "Open Ontario"
                         # "URL(.csv)", "URL(.zip)", "URL(.Rds)"
                         ),
             selected = "StatCan Table", inline=T) 



# h6("See Info for a sample of databases")
# h6("Search in [Search] Tab")

actionButton("run", "Run!", class = "btn-success",width = "100%")


```


```{r r.dt}

# . Select columns ----

renderUI({
  # req(r.dt())
  selectInput( "columns1", "Select columns :", r.dt() %>% names, 
               r.dt() %>% names,
               # c("Date", "val_norm", "Characteristics"),
               multiple = T)  
})

r.dt1 <- reactive({
  req(r.dt())
  # req(input$columns1)
  
    dt1 <- r.dt()  %>% .[, input$columns1, with=F ] 
  # r.dt()  %>% select(input$columns1)   
  # %>%
             # mutate_if(is.factor, as.character) 

  # fread(input$file1$datapath, stringsAsFactors = T) [ , ..input$columns1]
})






```

<!-- # LTE  -->

<!-- # Load (LTE) -->

# Loader





<!-- # Visualize  -->


## LEVEL 2 ........ {.tabset .tabset-fade }

### Load, Subset and Save


```{r}

# . Select a file ----

renderUI({
  
  switch(
    input$data.source, 
    
    "Upload file" = 
      fileInput(
        "file1",  h6("Upload CSV file:"), multiple = F, width="100%",
        accept = c("text/csv",
                   "text/comma-separated-values,text/plain",
                   ".csv")),
    
    "StatCan Table" = 
      textInput("data.id", "Enter StatCan table number:", "13-10-0810-01", width="100%"),
    
    "Open Ontario" = 
      textInput("data.id", "Enter Ontario table number:", "274b819c-5d69-4539-a4db-f2950794138c", width="100%"),
    
    textInput("data.id", "Enter URL for .csv, .xls or zip:", "https://health-infobase.canada.ca/src/data/covidLive/vaccination-coverage-byAgeAndSex-overTimeDownload.csv", width="100%")
  )
})


r.dt  <- reactive({

  switch (
    input$data.source,
          "uploaded" = {
            req(input$file1)
            fread(input$file1$datapath, stringsAsFactors = T)
          }
  )
  
  # dt <- cansim::get_cansim(input$stat.table) %>% setDT(dt)
  # dt[, Date := ymd(Date)]; 
  # 
  # dt$Date %>% max(na.rm = T)
  # dt$Date %>% min(na.rm = T)
  # dt %>% names()
  # dt$Date %>% unique()
  # dt$Characteristics %>% unique()
  # 
  # dt
})


renderDataTable({
  datatable.title( r.dt1() )
})

```


### Summarize

#### Stats
```{r}


renderPrint({  
  # summary( r.dt1()  ) 
  
  summary( r.dt1() %>%
             mutate_if(is.character, as.factor) %>%
             mutate_if(is.integer, as.factor)
  )
  
  
})
```

#### Structure
```{r}
renderPrint({  str(r.dt1())  })

  # dt$Date %>% max(na.rm = T)
  # dt$Date %>% min(na.rm = T)
  # dt %>% names()
  # dt$Date %>% unique()
  # dt$Characteristics %>% unique()
  

```




# Searcher 

## ...

### Search Open Canada databases by keyword

```{r r.dt0}


# textInput("stat.search", "Search keyword", "vital", width="100%")
# cat("Try: Vital, Covid, Hospital, Vaccin")
# 
# "Try: Vital, Covid, Hospital, Vaccin"
# 
# 
# checkboxInput("saveSearch", "Save search results in .csv file")
# 
# r.dt0  <- reactive({
#   
#   req(input$run)
#   dt0 <- search_cansim_cubes(input$stat.search) %>% setDT()
# 
#   
#   if (input$saveSearch)
#     fwrite(dt0, paste0("cansim-", input$stat.search, ".csv"))
#   
#   dt0
# })
# 
# renderDataTable({
#   datatable.title( r.dt0() )
# })




# ckanr_setup(url = "https://data.ontario.ca/")
# res <- resource_search(q = "name:Vaccin", as = "table") 
# res$count #7
# r <- res$results %>% setDT
# 
# 
# r <- res$results
# 
# 
# fwrite(res$results, "on-vaccine.csv", sep="\t")
# 
# res$results$package_id
# 
# res$results$id # = res$results$resource_id
# r$original_url
# r$name
# r$name_translated
# 
# r$description
# r$description_translated
# 
# r$created
# r$publically_available_date
# r$data_range_start
# r$data_range_end
# 
# 
# 
# 
# res <-resource_show(id ="274b819c-5d69-4539-a4db-f2950794138c")
# res$url
# destfile <- "Hospitalizations by vaccination status.csv"
# curl::curl_download(res$url, destfile)

```

# Visualizer


## .... Upper Row {.tabset .tabset-fade }

### Overview

### Multi-dimensional plot

<!-- ### Summary plot  -->

### Correlations plot

### Interactive time-series plot

### Interactive graph plot


# Info {data-icon="fa-info-circle"}


## .... Upper Row {.tabset .tabset-fade }


<!-- title: "Open Canada Vital Statistics Tracker" -->
<!-- subtitle: "Open Source Tool for Visualizing Open Data" -->


### About the App


#### Technical details



*Name: * Open Canada Vital Statistics Tracker   
*Type: * Web Application. Compatible with mobile devices   
*URL:* https://open-canada.github.io/Apps/vitals   
*Code Source:* https://github.com/open-canada/vitals     
*Data Source:* [Statistics Canada, Canadian Vital Statistics - Death database (CVSD)](https://www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey&SDDS=3233)   
*Licenses: * [Open Government license - Canada](http://open.canada.ca/en/open-government-licence-canada) (data), [MIT license](https://github.com/open-canada/vitals/blob/main/LICENSE) (code)


<!-- The code is under MIT license and is free to use by anyone.    -->
<!-- All the data is under [Open Government License - Canada](http://open.canada.ca/en/open-government-licence-canada). -->

<!-- Released: -->

<!-- For official information related to *Cases  following vaccination*,  see COVID-19 Daily Epidemiology Update at [https://health-infobase.canada.ca/covid-19/epidemiological-summary-covid-19-cases.html](https://health-infobase.canada.ca/covid-19/epidemiological-summary-covid-19-cases.html#a9). -->




<!-- It is free to use by anyone under the  MIT License.  -->

<!-- The Open Canada Vital Statistics Tracker  -->
<!-- #subtitle: "Open Source Tool for Visualizing Open Data" -->

#### Description


### Sample of Databases

#### StatCan CANSIM Tables


("Try: 13-10-0768-01, 1310076801, 13-10-0792-01")

Examples: 
"13-10-0768-01"
"13-10-0792-01"
"13-10-0783-01"

"13-10-0783-01"

"13-10-0427-01"
"13-10-0395-01"

"13-10-0415-01"
"13-10-0418-01"
"13-10-0415-01"

"13-26-0003" # Preliminary dataset on confirmed cases of COVID-19



#### Open Ontario API Tables



res <-resource_show(id ="274b819c-5d69-4539-a4db-f2950794138c")
res$url
destfile <- "Hospitalizations by vaccination status.csv"

#### Other from URL

##### csv 

https://health-infobase.canada.ca/src/data/covidLive/vaccination-coverage-byAgeAndSex-overTimeDownload.csv 

"https://github.com/open-canada/datasets/raw/main/statcan/13100810.csv"


##### 

https://www150.statcan.gc.ca/n1/tbl/csv/13100810-eng.zip 

https://github.com/open-canada/datasets/raw/main/statcan/13100810.Rds", 




```{r STOP_HERE, include=FALSE}
knitr::knit_exit()
```

